Diffantom: whole-brain diffusion MRI phantoms derived from real datasets of the \acrlong{hcp}

### 2015-09-09 paper/main.tex
latest changes, v2.3

Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the angular resolution in fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Digital simulations are increasingly popular because the complexity of whole-brain tractography
  can not be accounted for with current materials and proposed methodologies to build physical phantoms.
Early digital phantoms started with simulation of simple geometries
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate resolution of fiber crossings.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fiber crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic diffusion.
\cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.
An interesting outcome of \emph{FiberFox} is the phantom dataset\footnote{Available at
  \url{http://www.tractometer.org/ismrm_2015_challenge/}} created for the Tractography
  Challenge held in ISMRM 2015.
This dataset was derived from the tractography extracted in one \gls*{hcp} dataset.
In the tractogram, 25 fiber bundles of interest were manually segmented by experts.
Using \emph{FiberFox}, the segmentation of each bundle was mapped to an analytical
  description, and finally simulated the signal.

@@

\noindent\textbf{\textit{Microstructural model\textcolon}\label{sec:data_model}} %
The simulation process relies on a microstructural model derived from real data.
On one hand, the \emph{diffantomizer} workflow requires up to five fraction maps \tmaps{} of
  free- and hindered- diffusion (see \autoref{fig:figure1}, box A).
These compartments will be derived from the macroscopic structure of tissues within the brain,
  specified in the following order\footnote{Corresponding to the \emph{5TT format} established
  with the latest version 3.0 of \emph{MRTrix} \citep{tournier_mrtrix_2012}}:
  \gls*{cgm}, \gls*{dgm}, \gls*{wm}, \gls*{csf}, and abnormal tissue\footnote{Since here we
  simulate healthy subjects, the last fraction map $T_5$ is empty and can be omitted}.
On the other hand, the restricted-diffusion compartments are specified by up to three volume fractions \fmaps{}
  of three single fiber populations per voxel along with their corresponding direction maps \vmaps{}.

The process to obtain the microstructural model from one dataset of the \gls*{hcp} can be described
  as follows (see also \autoref{fig:figure1}, box B):
1) The fiber orientation maps $\{\mathbf{V}_i\}$ and their corresponding estimations of volume fraction $\{F'_i\}$ are
  obtained using the ball-and-stick model for multi-shell data of \gls*{bedpostx}
  on the \gls*{dmri} data.
The \gls*{hcp} recommends \gls*{bedpostx} to reconstruct their data \citep{glasser_minimal_2013}.
A further advantage is that \gls*{bedpostx} exploits the multi-shell acquisitions of the \gls*{hcp} while
  operating at whole-brain level.
2) A \gls*{fa} map is obtained after fitting a tensor model with \emph{MRTrix}.
As we shall see in the Appendix, the \gls*{fa} is used to infer $F_1$ (the fraction map of the most prevalent fiber),
  avoiding the extremely noisy estimation of $F'_1$ performed by \gls*{bedpostx} in the previous step.
3) The original fiber fractions $\{F'_i\}$ and the \gls*{fa} map are denoised with a nonlocal means filter included
  in \emph{dipy} \citep{garyfallidis_dipy_2014}.
This step produces an important smoothing of the maps, while preserving the edges.
Smoothing is also beneficial in simplifying the voxel-wise diffusion model.
4) The macrostructural fractions $\{T'_j\}$ are extracted from the \acrlong*{t1} image of the dataset,
  using standard \emph{FSL} segmentation tools \citep{jenkinson_fsl_2012}.
5) The images obtained previously (\gls*{fa} map, $\{\mathbf{V}_i\}$, $\{F'_i\}$, and $\{T'_j\}$)
  are combined as described in the \nameref{sec:appendix} to generate the final microstructural model
  ($\{\mathbf{V}_i\}$, $\{F_i\}$, and $\{T_j\}$), presented in \autoref{fig:figure1}-A.

@@

\noindent\textbf{\textit{Interpretation and recommended uses\textcolon}\label{sec:data_use}} %
To illustrate the features of \emph{Diffantom}, the example dataset underwent a simplified
  connectivity pipeline including \gls*{csd} and probabilistic tractography from
  \emph{MRTrix} \citep{tournier_mrtrix_2012}.
\Gls*{csd} was reconstructed using 8$^\text{th}$-order spherical harmonics, and tractography with 1.6\e{6}
  seed points evenly distributed across a dilated mask of the \gls*{wm} tissue.
\autoref{fig:figure2}, box A1, shows the result of the tractography obtained with such pipeline.
Finally, we applied \emph{tract querier} \citep{wassermann_on_2013} on segmenting important fiber bundles such
  as the \gls*{cst} and the forceps minor (see \autoref{fig:figure2}, box A2).
Particularly, due to its location nearby the orbitofrontal lobe, the forceps minor is generally affected by
  susceptibility distortions.

We recommend \emph{Diffantom} as ground-truth in verification and validation frameworks
  (\autoref{fig:figure2}, box B) for testing pipelines.
\emph{Diffantom} is applicable in the unit testing of algorithms, the integration testing of
  modules in workflows, and the overall system testing.
Some potential applications follow:
\begin{itemize}
  \item Investigating the impact of different diffusion sampling schemes on the local microstructure
    model of choice and on the subsequent global tractography outcome.
  Since the gradient scheme can be set by the user, \emph{Diffantom} can be seen as a mean to translate the so-called
  \emph{b-matrix} of the source dataset to any target scheme.
  \item Assessment of sensitivity and robustness to imaging artifacts (noise, \acrlong{pve} and \gls*{csf} contamination,
    susceptibility-derived warping, Eddy-currents-derived distortions, etc.) at unit, integration and systems testing levels.
  \item Using \emph{Diffantom} as in box B of \autoref{fig:figure2}, it is possible to apply binary classification measures
    to evaluate the resulting connectivity matrix.
  Considering the connectivity matrix of the \emph{reference Diffantom} and the resulting
    matrix of the \emph{test Diffantom}, the \acrfull*{roc} of the pipeline can be characterized.
  \item Simulation of pathological brains by altering the microstructural
    model accordingly (e.g. as tumors were simulated in \cite{kaus_simulation_2000}).
\end{itemize}
In order to exemplify one of these intended uses, we also release a \emph{Diffantom} including the susceptibility-derived
  distortion in simulation.

@@

\begin{figure}[h!]
\begin{center}
\includestandalone[width=\linewidth]{figure01}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure1} Figure \arabic{figure}. }{%
\textbf{A. Microstructural model of \emph{Diffantom}}.
The phantom is simulated from an underlying microstructural model specified with the
  following volume-fraction maps: three hindered-diffusion compartments $\{T_1, T_2, T_3\}$,
  one free-diffusion compartment $T_4$ corresponding to the \gls*{csf},
  three restricted-diffusion compartments $\{F_i\}$, and three vectorial maps
  associated with the local fiber directions $\{\mathbf{V}_i\}$.
Please note the piece-wise linear function of the color scale to enable visibility of small volume fractions.
\textbf{B. The \emph{diffantomizer} workflow, a workflow to generate \emph{diffantoms}}.
The pipeline to generate phantoms from any \gls*{hcp} dataset is presented in the lower box.
Once the microstructural model shown in the upper box has been prepared as described in \nameref{sec:data_model},
  the local orientations are computed and fed into \emph{Phantomas} to finally simulate the signal.
}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includestandalone[width=\linewidth]{figure02}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure2} Figure \arabic{figure}. }{%
\textbf{A1. Example dataset.} The tractogram of fibers crossing slice 56 of \emph{Diffantom} as
  extracted with \emph{MRTrix} is presented, represented over the corresponding slice of the
  \lowb{} volume.
\textbf{A2. Segmentation of some fiber bundles in the example dataset.} We used \emph{tract\_querier} to
  segment the right \gls*{cst} represented with blue streamlines, the left \gls*{cst} (red streamlines),
  and the forceps minor (green streamlines).
  The box includes the slice 56 of the \lowb{} and the pial surface is represented with transparency.
\textbf{B. Recommended use of \emph{Diffantom}}.
The phantom is designed to be used as ground-truth information in evaluation frameworks,
  to implement unit test of algorithms, to check integration of processing units within
  pipelines or to validate complete workflows.
For instance, in order to evaluate artifacts, a perturbation can be induced in the microstructural
  model or after simulation to provide reference and test datasets.
}
\end{figure}

>>>>

Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the angular resolution in fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Digital simulations are increasingly popular because the complexity of whole-brain tractography
  can not be accounted for with current materials and proposed methodologies to build physical phantoms.
Early digital phantoms started with simulation of simple geometries
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate the angular resolution as well.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fiber crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic diffusion.
\cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.
An interesting outcome of \emph{FiberFox} is the phantom dataset\footnote{Available at
  \url{http://www.tractometer.org/ismrm_2015_challenge/}} created for the Tractography
  Challenge held in ISMRM 2015.
This dataset was derived from the tractography extracted in one \gls*{hcp} dataset.
In the tractogram, 25 fiber bundles of interest were manually segmented by experts.
Using \emph{FiberFox}, the segmentation of each bundle was mapped to an analytical
  description, and finally simulated the signal.

@@

\noindent\textbf{\textit{Microstructural model\textcolon}\label{sec:data_model}} %
The simulation process relies on a microstructural model derived from real data.
On one hand, the \emph{diffantomizer} workflow requires up to five fraction maps \tmaps{} of
  free- and hindered- diffusion (see \autoref{fig:figure1}, panel A).
These compartments will be derived from the macroscopic structure of tissues within the brain,
  specified in the following order\footnote{Corresponding to the \emph{5TT format} established
  with the latest version 3.0 of \emph{MRTrix} \citep{tournier_mrtrix_2012}}:
  \gls*{cgm}, \gls*{dgm}, \gls*{wm}, \gls*{csf}, and abnormal tissue\footnote{Since here we
  simulate healthy subjects, the last fraction map $T_5$ is empty and can be omitted}.
On the other hand, the restricted-diffusion compartments are specified by up to three volume fractions \fmaps{}
  of three single fiber populations per voxel along with their corresponding direction maps \vmaps{}.

The process to obtain the microstructural model from one dataset of the \gls*{hcp} can be described
  as follows (see also \autoref{fig:figure1}, panel B):
1) The fiber orientation maps $\{\mathbf{V}_i\}$ and their corresponding estimations of volume fraction $\{F'_i\}$ are
  obtained using the ball-and-stick model for multi-shell data of \gls*{bedpostx}
  on the \gls*{dmri} data.
The \gls*{hcp} recommends \gls*{bedpostx} to reconstruct their data \citep{glasser_minimal_2013}.
A further advantage is that \gls*{bedpostx} exploits the multi-shell acquisitions of the \gls*{hcp} while
  operating at whole-brain level.
2) A \gls*{fa} map is obtained after fitting a tensor model with \emph{MRTrix}.
As we shall see in the Appendix, the \gls*{fa} is used to infer $F_1$ (the fraction map of the most prevalent fiber),
  avoiding the extremely noisy estimation of $F'_1$ performed by \gls*{bedpostx} in the previous step.
3) The original fiber fractions $\{F'_i\}$ and the \gls*{fa} map are denoised with a nonlocal means filter included
  in \emph{dipy} \citep{garyfallidis_dipy_2014}.
This step produces an important smoothing of the maps, while preserving the edges.
Smoothing is also beneficial in simplifying the voxel-wise diffusion model.
4) The macrostructural fractions $\{T'_j\}$ are extracted from the \acrlong*{t1} image of the dataset,
  using standard \emph{FSL} segmentation tools \citep{jenkinson_fsl_2012}.
5) The images obtained previously (\gls*{fa} map, $\{\mathbf{V}_i\}$, $\{F'_i\}$, and $\{T'_j\}$)
  are combined as described in the \nameref{sec:appendix} to generate the final microstructural model
  ($\{\mathbf{V}_i\}$, $\{F_i\}$, and $\{T_j\}$), presented in \autoref{fig:figure1}-A.

@@

\noindent\textbf{\textit{Interpretation and recommended uses\textcolon}\label{sec:data_use}} %
To illustrate the features of \emph{Diffantom}, the example dataset underwent a simplified
  connectivity pipeline including \gls*{csd} and probabilistic tractography from
  \emph{MRTrix} \citep{tournier_mrtrix_2012}.
\Gls*{csd} was reconstructed using 8$^\text{th}$-order spherical harmonics, and tractography with 1.6\e{6}
  seed points evenly distributed across a dilated mask of the \gls*{wm} tissue.
\autoref{fig:figure2}, panels A1 and A3, show the result of the tractography obtained with such pipeline for
  the original \emph{Diffantom} and a distorted version.
Finally, we applied \emph{tract querier} \citep{wassermann_on_2013} on segmenting important fiber bundles such
  as the \gls*{cst} and the forceps minor (see \autoref{fig:figure2}, panels A2, A4).
Particularly, due to its location nearby the orbitofrontal lobe, the forceps minor is generally affected by
  susceptibility distortions.

We recommend \emph{Diffantom} as ground-truth in verification and validation frameworks
  (\autoref{fig:figure2}, panel B) for testing pipelines.
\emph{Diffantom} is applicable in the unit testing of algorithms, the integration testing of
  modules in workflows, and the overall system testing.
Some potential applications follow:
\begin{itemize}
  \item Investigating the impact of different diffusion sampling schemes on the local microstructure
    model of choice and on the subsequent global tractography outcome.
  Since the gradient scheme can be set by the user, \emph{Diffantom} can be seen as a mean to translate the so-called
  \emph{b-matrix} of the source dataset to any target scheme.
  \item Assessment of sensitivity and robustness to imaging artifacts (noise, \acrlong{pve} and \gls*{csf} contamination,
    susceptibility-derived warping, Eddy-currents-derived distortions, etc.) at unit, integration and systems testing levels.
  \item Using \emph{Diffantom} as in panel B of \autoref{fig:figure2}, it is possible to apply binary classification measures
    to evaluate the resulting connectivity matrix.
  Considering the connectivity matrix of the \emph{reference Diffantom} and the resulting
    matrix of the \emph{test Diffantom}, the \acrfull*{roc} of the pipeline can be characterized.
  \item Simulation of pathological brains by altering the microstructural
    model accordingly (e.g. as tumors were simulated in \cite{kaus_simulation_2000}).
\end{itemize}
In order to exemplify one of these intended uses, we also release a \emph{Diffantom} including the susceptibility-derived
  distortion in simulation.

@@

\begin{figure}[h!]
\begin{center}
\includestandalone[width=\linewidth]{figure01}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure1} Figure \arabic{figure}. }{%
\textbf{A. Microstructural model of \emph{Diffantom}}.
The phantom is simulated from an underlying microstructural model specified with the
  following volume-fraction maps: three hindered-diffusion compartments $\{T_1, T_2, T_3\}$,
  one free-diffusion compartment $T_4$ corresponding to the \gls*{csf},
  three restricted-diffusion compartments $\{F_i\}$, and three vectorial maps
  associated with the local fiber directions $\{\mathbf{V}_i\}$.
Please note the piece-wise linear function of the color scale to enable visibility of small volume fractions.
\textbf{B. The \emph{diffantomizer} workflow, a workflow to generate \emph{diffantoms}}.
The pipeline to generate phantoms from any \gls*{hcp} dataset is presented in the lower panel.
Once the microstructural model shown in the upper panel has been prepared as described in \nameref{sec:data_model},
  the local orientations are computed and fed into \emph{Phantomas} to finally simulate the signal.
}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includestandalone[width=\linewidth]{figure02}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:figure2} Figure \arabic{figure}. }{%
\textbf{A. Example dataset.}
A1 and A3 show the tractogram of fibers crossing slice 56 of \emph{Diffantom} as
  extracted with \emph{MRTrix}, represented over the corresponding slice of the
  \lowb{} volume for the original (A1) and the distorted (A3) phantoms, with a gray
  frame highlighting the absence of important tracks.
Panels A2 and A4 show the segmentation of the right \gls*{cst} represented with blue
  streamlines, the left \gls*{cst} (red streamlines), and the forceps minor (green streamlines)
  using \emph{tract\_querier}.
A2 and A4 include the slice 56 of the \lowb{} and the pial surface is represented
  with transparency.
In the distorted \emph{Diffantom} (A4) the forceps minor was not detected.
\textbf{B. Recommended use of \emph{Diffantom}}.
The phantom is designed to be used as ground-truth information in evaluation frameworks,
  to implement unit test of algorithms, to check integration of processing units within
  pipelines or to validate complete workflows.
For instance, in order to evaluate artifacts, a perturbation can be induced in the microstructural
  model or after simulation to provide reference and test datasets.
}
\end{figure}
